\texttt{treats}: a modular \texttt{R} package for simulating trees and traits.

### 2023-11-23 inst/treats_paper.Rmd
CRAN submission fixes

3. Here I present the `treats` package, a modular `R` package for trees and traits simulations.
This package is based on a simple birth death algorithm from which all steps can easily be modified by users.
It also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
It also comes with an extend manual regularly updated following users' questions or suggestions.

@@

The `treats` package comes with internal documentation (e.g. `?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html) ([http://tguillerme.github.io/treats.html](http://tguillerme.github.io/treats.html).
This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].

>>>>

3. Here I present the `treats` package, a modular `R` package for trees and traits simulations.
This package is based on a simple birth death algorithm from which all steps can easily be modified by users.

4. `treats` also provides a tidy interface through the `treats` object, allowing users to easily run reproducible simulations.
It also comes with an extend manual regularly updated following users' questions or suggestions.

@@

The `treats` package comes with internal documentation (e.g. `?treats`) but also with a thorough and extended vignette in a gitbook format: the [`treats` manual](http://tguillerme.github.io/treats.html).
This manual is designed so that it can be regularly updated and enhanced through the lifetime of the package facilitating the interface between methods development and usage [@cooper2016dark].

### 2023-11-23 paper/treats_paper_anonymous_lineno_nopath.tex
CRAN submission fixes

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Thomas Guillerme\(^1\)}
\date{2023-11-22}

@@

\(^1\) School of Biosciences, University of Sheffield, Sheffield, S10
2TN, United Kingdom.
\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Simulating biological realistic data is an important step to
  understand and investigate biodiversity. Simulated data can be used to
  generate null, base line or neutral models. These can be used either
  in comparison to observed data to estimate the mechanisms that
  generated the data. Or they can be used to explore, understand and
  develop theoretical advances by proposing toy models.
\item
  In evolutionary biology, simulations often involve the need of an
  evolutionary process where descent with modification is at the core of
  how the simulated data is generated. These evolutionary processes can
  then be nearly infinitely modified to include complex processes that
  affect the simulations such as traits co-evolution, competition
  mechanisms or mass extinction events.
\item
  Here I present the \texttt{treats} package, a modular \texttt{R}
  package for trees and traits simulations. This package is based on a
  simple birth death algorithm from which all steps can easily be
  modified by users. It also provides a tidy interface through the
  \texttt{treats} object, allowing users to easily run reproducible
  simulations. It also comes with an extend manual regularly updated
  following users' questions or suggestions.
\end{enumerate}

@@

\begin{figure}
\hypertarget{figure1}{%
\centering
\includegraphics{../inst/gitbook/treats_structure.png}
\caption{\texttt{treats} package workflow: the \texttt{treats} algorithm
generates a tree and traits using inbuilt \texttt{"traits"} objects that
contain the instructions on how to generate the trait data (e.g.~which
process? how many dimensions?); \texttt{"modifiers"} objects that
contains instructions on how to ``grow'' the tree (e.g.~by linking
speciation to trait values, or to the current number of species); and
\texttt{"events"} objects that can modify the tree structure,
\texttt{"modifiers"} or \texttt{"traits"} depending on specific
conditions (e.g.~80\% of species with positive trait values go extinct
after reaching a specific time).}\label{figure1}
}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

For this event, the target is the number of taxa in the simulations.
This is indicated using the \texttt{target\ =\ "taxa"} argument. Then
the event is triggered using the argument \texttt{time.condition(15)}
and modifies the internal \texttt{lineage} object using the
\texttt{trait.extinction(x\ =\ 0,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{})}
function. \texttt{time.condition} and \texttt{trait.extinction} are both
function factories which are not gonna be detailed here (see Wickham
(2019)). Effectively these arguments can be passed directly as standard
functions. For example, to trigger the event when reaching time 15 we
can use the following function:

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/prepdata-1.pdf}
\caption{Observed phylogeny (A) and disparity through time (B) in the
subset of Beck and Lee (2014)'s dataset. The red line represents the
K-Pg boundary. Is this change in disparity related to the K-Pg mass
extinction? Or, more generally, is it even possible to detect potential
changes of disparity due to a mass extinction event?}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtree-1.pdf}
\caption{A randomly simulated tree with similar properties as the
observed one. Note that the time here is expressed in arbitrary units.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraits-1.pdf}
\caption{A randomly simulated tree with a 2 dimensional random Brownian
Motion trait. Orange, light blue and dark blue dots respectively
represent nodes, fossils and living species.}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a random mass extinction}
\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition =} \KeywordTok{time.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simtreetraitsevents-1.pdf}
\caption{Two different mass extinction events: (A) 75\% of species go
extinct; (B) all species with positive trait values go extinct. The red
line marks the extinction event.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{treats_paper_files/figure-latex/simdisparity-1.pdf}
\caption{Comparison of the observed sum of variances (dashed line) to
the simulated one (ligth and dark grey polygons and black line
representing respectively the 95\% and 50\% confidence interval and the
median value).}
\end{figure}

@@

The \texttt{treats} package comes with internal documentation
(e.g.~\texttt{?treats}) but also with a thorough and extended vignette
in a gitbook format: the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}
(\url{http://tguillerme.github.io/treats.html}. This manual is designed
so that it can be regularly updated and enhanced through the lifetime of
the package facilitating the interface between methods development and
usage (Cooper, Thomas, and FitzJohn 2016).

@@

The \texttt{treats} package is available on
\href{https://github.com/TGuillerme/treats}{GitHub} with more associated
information. All the versions of the package are archived on ZENODO with
associated
\href{https://doi.org/10.5281/zenodo.7970384}{DOI:10.5281/zenodo.7970384}.

>>>>

\title{\texttt{treats}: a modular \texttt{R} package for simulating
trees and traits.}
\author{Anonymous\(^1\)}% \author{Thomas Guillerme\(^1\)}
\date{2023-11-23}

@@

\(^1\) Anonymous % \(^1\) School of Biosciences, University of Sheffield, Sheffield, S10 2TN, United Kingdom.
\href{mailto:anonymous}{\nolinkurl{anonymous}}% \href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Simulating biological realistic data is an important step to
  understand and investigate biodiversity. Simulated data can be used to
  generate null, base line or neutral models. These can be used either
  in comparison to observed data to estimate the mechanisms that
  generated the data. Or they can be used to explore, understand and
  develop theoretical advances by proposing toy models.
\item
  In evolutionary biology, simulations often involve the need of an
  evolutionary process where descent with modification is at the core of
  how the simulated data is generated. These evolutionary processes can
  then be nearly infinitely modified to include complex processes that
  affect the simulations such as traits co-evolution, competition
  mechanisms or mass extinction events.
\item
  Here I present the \texttt{treats} package, a modular \texttt{R}
  package for trees and traits simulations. This package is based on a
  simple birth death algorithm from which all steps can easily be
  modified by users.
\item
  \texttt{treats} also provides a tidy interface through the
  \texttt{treats} object, allowing users to easily run reproducible
  simulations. It also comes with an extend manual regularly updated
  following users' questions or suggestions.
\end{enumerate}

@@

\begin{figure}
\hypertarget{figure1}{%
\centering
\includegraphics{treats_structure.png}
\caption{\texttt{treats} package workflow: the \texttt{treats} algorithm
generates a tree and traits using inbuilt \texttt{"traits"} objects that
contain the instructions on how to generate the trait data (e.g.~which
process? how many dimensions?); \texttt{"modifiers"} objects that
contains instructions on how to ``grow'' the tree (e.g.~by linking
speciation to trait values, or to the current number of species); and
\texttt{"events"} objects that can modify the tree structure,
\texttt{"modifiers"} or \texttt{"traits"} depending on specific
conditions (e.g.~80\% of species with positive trait values go extinct
after reaching a specific time).}\label{figure1}
}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

For this event, the target is the number of taxa in the simulations.
This is indicated using the \texttt{target\ =\ "taxa"} argument. Then
the event is triggered using the argument \texttt{age.condition(15)} and
modifies the internal \texttt{lineage} object using the
\texttt{trait.extinction(x\ =\ 0,\ condition\ =\ \textasciigrave{}\textgreater{}=\textasciigrave{})}
function. \texttt{age.condition} and \texttt{trait.extinction} are both
function factories which are not gonna be detailed here (see Wickham
(2019)). Effectively these arguments can be passed directly as standard
functions. For example, to trigger the event when reaching time 15 we
can use the following function:

@@

\begin{figure}
\centering
\includegraphics{prepdata-1.pdf}
\caption{Observed phylogeny (A) and disparity through time (B) in the
subset of Beck and Lee (2014)'s dataset. The red line represents the
K-Pg boundary. Is this change in disparity related to the K-Pg mass
extinction? Or, more generally, is it even possible to detect potential
changes of disparity due to a mass extinction event?}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{simtree-1.pdf}
\caption{A randomly simulated tree with similar properties as the
observed one. Note that the time here is expressed in arbitrary units.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{simtreetraits-1.pdf}
\caption{A randomly simulated tree with a 2 dimensional random Brownian
Motion trait. Orange, light blue and dark blue dots respectively
represent nodes, fossils and living species.}
\end{figure}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#\# Creating a random mass extinction}
\NormalTok{random\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target       =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition    =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{random.extinction}\NormalTok{(}\FloatTok{0.75}\NormalTok{))}
\CommentTok{\#\# Creating an extinction that removes species with positive trait values}
\NormalTok{positive\_extinction \textless{}{-}}\StringTok{ }\KeywordTok{make.events}\NormalTok{(}
    \DataTypeTok{target =} \StringTok{"taxa"}\NormalTok{,}
    \DataTypeTok{condition =} \KeywordTok{age.condition}\NormalTok{(}\DecValTok{15}\NormalTok{),}
    \DataTypeTok{modification =} \KeywordTok{trait.extinction}\NormalTok{(}\DataTypeTok{x =} \DecValTok{0}\NormalTok{, }\DataTypeTok{condition =} \StringTok{\textasciigrave{}}\DataTypeTok{\textgreater{}=}\StringTok{\textasciigrave{}}\NormalTok{))}

@@

\begin{figure}
\centering
\includegraphics{simtreetraitsevents-1.pdf}
\caption{Two different mass extinction events: (A) 75\% of species go
extinct; (B) all species with positive trait values go extinct. The red
line marks the extinction event.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{simdisparity-1.pdf}
\caption{Comparison of the observed sum of variances (dashed line) to
the simulated one (ligth and dark grey polygons and black line
representing respectively the 95\% and 50\% confidence interval and the
median value).}
\end{figure}

@@

The \texttt{treats} package comes with internal documentation
(e.g.~\texttt{?treats}) but also with a thorough and extended vignette
in a gitbook format: the
\href{http://tguillerme.github.io/treats.html}{\texttt{treats} manual}
(\href{https://anonymous.4open.science/r/treats-006B}{http://tguillerme.github.io/treats.html}).
This manual is designed so that it can be regularly updated and enhanced
through the lifetime of the package facilitating the interface between
methods development and usage (Cooper, Thomas, and FitzJohn 2016).

@@

The \texttt{treats} package is available on
\href{https://anonymous.4open.science/r/treats-006B}{GitHub} with more
associated information. All the versions of the package are archived on
ZENODO with associated
\href{https://doi.org/10.5281/zenodo.7970384}{DOI:10.5281/zenodo.7970384}.
